D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-06 ootavacuous.tex
Small changes for D.1 and D.2
Here is a set of small edits to sections D.1 and D.2.  They are mostly
just stylistic improvements.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Because both \co{x} and \co{y} are nonvolatile, the compiler is
free to assume that the value loaded by line~4 will be some value
stored to \co{x} or its initial value.
Because there are no stores to \co{x}, the only remaining possibility
is the default-initialized value of zero.
Therefore, the C++ compiler is within its rights to substitute
the constant zero for \co{r1} on line~5, eliminating the semantic
dependency that would otherwise extend from line~4 to line~5.

@@

In this case, the \co{volatile} keyword alerts the compiler to
the possibility of interfering accesses to \co{x} from outside the
program, for example, due to \co{x} being:

\begin{itemize}
\item	Allocated in an MMIO region of the address space.
\item	Modified by an unknown-to-the-compiler thread
	or signal handler.
\item	Subject to I/O-device DMA operations.
\item	Modified by debugger commands.
\item	Modified by accesses to facilities such as \path{/dev/mem}.
\item	Modified by as-yet-unwritten dynamically linked libraries.
\item	Modified by some other mechanism of the reader's choosing.
\end{itemize}

These possibilities force the compiler to preserve the semantic
dependency from line~4 to line~5.
This preserved dependency helps prevent the formation of OOTA cycles
even in situations where there is no interference.\footnote{
	When there is interference, the definition of ``OOTA cycle''
	does not apply.}

@@

Starting with the intuitive execution where the values loaded from \co{x}
and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
that differ only in the value loaded from either \co{x} or \co{y} will
still affect the value stored to \co{z} by line~11.
There is therefore one semantic dependency from the load from \co{x}
to the store to \co{z} and another separate semantic dependency from
the load from \co{y} to the store to \co{z}, which matches typical
definitions of semantic dependency.
Yet if the initial values were instead obtained as input,
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
could be thought of as being two different executions of the exact same
program.

@@

One might ask why semantic dependencies can be nontrivial when hardware
address, control, and data dependencies are much more straightforward.
The reason is that, unlike compilers, hardware:
(1)~Respects dependencies even when they reduce to a constant,
(2)~Less-aggressively optimizes, and
(3)~Lacks undefined behavior.

This situation further underscores the per-execution nature of semantic
dependencies as well as the need for an improved definition of ``semantic
dependency''.
In
Section~\ref{sec:A Definition of Semantic Dependency},
we expand the traditional definition to include dependencies extending
from groups of loads to a single store.

>>>>

Because both \co{x} and \co{y} are nonvolatile, a C++ compiler is
free to assume that the value loaded by line~4 will be either some value
stored to \co{x} or its initial value.
And because there are no stores to \co{x}, the only remaining possibility
is the default-initialized value of zero.
Therefore, the compiler is within its rights to substitute
the constant zero for \co{r1} on line~5, eliminating the semantic
dependency that would otherwise extend from line~4 to line~5.

@@

In this case, the \co{volatile} keyword alerts the compiler to
the possibility of interfering changes to \co{x} from outside the
program, for example, due to \co{x} being:
\begin{itemize}
\item	Allocated in an MMIO region of the address space.
\item	Modified by an unknown-to-the-compiler thread
	or signal handler.
\item	Subject to I/O-device DMA operations.
\item	Modified by debugger commands.
\item	Modified by accesses to facilities such as \path{/dev/mem}.
\item	Modified by as-yet-unwritten dynamically linked libraries.
\item	Modified by some other mechanism of the reader's choosing.
\end{itemize}
These possibilities force the compiler to preserve the semantic
dependency from line~4 to line~5,
and this preserved dependency helps prevent the formation of OOTA cycles
in environments where there is no interference.\footnote{
	When there is interference, the definition of ``OOTA cycle''
	does not apply.}

@@

Starting with the intuitive execution where the values loaded from \co{x}
and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
that differ only in the value loaded from either \co{x} or \co{y} will
now affect the value stored to \co{z} by line~11.
There is therefore one semantic dependency from the load from \co{x}
to the store to \co{z} and another separate semantic dependency from
the load from \co{y} to the store to \co{z}, which matches typical
definitions of semantic dependency.
Yet if the initial values were instead obtained as input,
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
could be thought of as being two different executions of the exact same
program.

@@

One might ask why semantic dependencies can be so complicated when hardware
dependencies are much more straightforward.
The reason is that, unlike compilers, hardware:
\begin{itemize}
\item	Respects dependencies even when they reduce to a constant,
\item	Optimizes much less aggressively, if at all, and
\item	Lacks undefined behavior.
\end{itemize}
This situation further underscores the per-execution nature of semantic
dependencies as well as the need for an improved definition of ``semantic
dependency''.
In
Section~\ref{sec:A Definition of Semantic Dependency},
we expand the traditional definition to include dependencies extending
from groups of loads to a single store.
